---
title: "Publication-ready volcano plots with enhanced colouring and labeling"
author: "Kevin Blighe"
date: "`r Sys.Date()`"
package: "`r packageVersion('EnhancedVolcano')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: united
    highlight: tango
fig_width: 7
bibliography: library.bib
vignette: >
    %\VignetteIndexEntry{Publication-ready volcano plots with enhanced colouring and labeling}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\usepackage[utf8]{inputenc}
---

# Introduction

Volcano plots represent a useful way to visualise the results of differential
expression analyses. Here, we present a highly-configurable function that
produces publication-ready volcano plots. EnhancedVolcano [@EnhancedVolcano]
will attempt to fit as many transcript names in the plot window as possible,
thus avoiding 'clogging' up the plot with labels that could not otherwise
have been read. Other functionality allows the user to identify up to 3
different types of attributes in the same plot space via colour, shape, and
shade parameter configurations.


```{r, echo=FALSE}

  library(knitr)
  opts_chunk$set(tidy = FALSE, message = FALSE, warning = FALSE)

```

# Installation

## 1. Download the package from Bioconductor

```{r getPackage, eval=FALSE}

  if (!requireNamespace('BiocManager', quietly = TRUE))
    install.packages('BiocManager')
    BiocManager::install('EnhancedVolcano')

```

Note: to install development version:

```{r getPackageDevel, eval=FALSE}

  devtools::install_github('kevinblighe/EnhancedVolcano')

```

## 2. Load the package into R session

```{r Load, message=FALSE}

  library(EnhancedVolcano)

```


# Quick start

For this example, we will follow the tutorial (from Section 3.1) of [RNA-seq workflow: gene-level
exploratory analysis and differential expression](http://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html). Specifically, we will load the 'airway' data, where different airway smooth muscle cells were treated with dexamethasone.

```{r}

  library(airway)
  library(magrittr)

  data('airway')
  airway$dex %<>% relevel('untrt')

```

Conduct differential expression using DESeq2 in order to create 2 sets of results:

```{r}

  library('DESeq2')

  dds <- DESeqDataSet(airway, design = ~ cell + dex)
  dds <- DESeq(dds, betaPrior=FALSE)
  res1 <- results(dds,
    contrast = c('dex','trt','untrt'))
  res1 <- lfcShrink(dds,
    contrast = c('dex','trt','untrt'), res=res1)
  res2 <- results(dds,
    contrast = c('cell', 'N061011', 'N61311'))
  res2 <- lfcShrink(dds,
    contrast = c('cell', 'N061011', 'N61311'), res=res2)

```

## Plot the most basic volcano plot

For the most basic volcano plot, only a single data-frame or -matrix of
test results is required, containing transcript names, log2FC, and adjusted
or unadjusted P values. The default cut-off for log2FC is >|2|; the default
cut-off for P value is 10e-6.

```{r ex1, fig.height = 8.5, fig.width = 7, fig.cap = "Plot the most basic volcano plot."}

  EnhancedVolcano(res1,
    lab = rownames(res1),
    x = 'log2FoldChange',
    y = 'pvalue',
    xlim = c(-5, 8))

```

# Advanced features

Virtually all aspects of an EnhancedVolcano plot can be configured for the
purposes of accommodating all types of statistical distributions and labelling
preferences. By default, EnhancedVolcano will only attempt to label genes that
pass the thresholds that you set for statistical significance, i.e., 'pCutoff' and
'FCcutoff'. In addition, it will only label as many of these that can reasonably
fit in the plot space. The user can optionally supply a vector of transcript names
(as 'selectLab') that s/he wishes to label in the plot.

## Modify cut-offs for log2FC and P value; specify title; adjust point and label size

The default P value cut-off of 10e-6 may be too relaxed for most studies, which
may therefore necessitate increasing this threshold by a few orders of magnitude.
Equally, the log2FC cut-offs may be too stringent, given that moderated 'shrunk'
estimates of log2FC differences in differential expression analysis can now be
calculated.

In this example, we also modify the point and label size, which can help to
improve clarity where many transcripts went into the differential expression analysis.

```{r ex2, fig.height = 8.5, fig.width = 7, fig.cap = "Modify cut-offs for log2FC and P value; specify title; adjust point and label size."}

  EnhancedVolcano(res2,
    lab = rownames(res2),
    x = 'log2FoldChange',
    y = 'pvalue',
    xlim = c(-8, 8),
    title = 'N061011 versus N61311',
    pCutoff = 10e-16,
    FCcutoff = 1.5,
    transcriptPointSize = 1.5,
    transcriptLabSize = 3.0)

```

## Adjust colour and alpha for point shading

The default colour scheme may not be to everyone's taste. Here we make it such that
only the transcripts passing both the log2FC and P value thresholds are coloured red,
with everything else black. We also adjust the value for 'alpha', which controls the
transparency of the plotted points: 1 = 100% opaque; 0 = 100% transparent.

```{r ex3, fig.height = 8.5, fig.width = 7, fig.cap = "Adjust colour and alpha for point shading."}

  EnhancedVolcano(res2,
    lab = rownames(res2),
    x = 'log2FoldChange',
    y = 'pvalue',
    xlim = c(-8, 8),
    title = 'N061011 versus N61311',
    pCutoff = 10e-16,
    FCcutoff = 1.5,
    transcriptPointSize = 1.5,
    transcriptLabSize = 3.0,
    col=c('black', 'black', 'black', 'red3'),
    colAlpha = 1)

```

## Adjust shape of plotted points

It can help, visually, to also plot different points as different shapes. The
default shape is a circle. The user can specify their own shape encoding via
the 'shape' parameter, which accepts either a single or four possible values:
if four values, these then map to the standard designation that is also
assigned by the colours; if a single value, all points are shaped with this value.

For more information on shape encoding search online at
[ggplot2 Quick Reference: shape](http://sape.inf.usi.ch/quick-reference/ggplot2/shape)

```{r ex4, fig.height = 8.5, fig.width = 7, fig.cap = "Adjust shape of plotted points."}

 EnhancedVolcano(res2,
    lab = rownames(res2),
    x = 'log2FoldChange',
    y = 'pvalue',
    xlim = c(-8, 8),
    title = 'N061011 versus N61311',
    pCutoff = 10e-16,
    FCcutoff = 1.5,
    transcriptPointSize = 3.0,
    transcriptLabSize = 3.0,
    shape = 8,
    colAlpha = 1)

  EnhancedVolcano(res2,
    lab = rownames(res2),
    x = 'log2FoldChange',
    y = 'pvalue',
    xlim = c(-8, 8),
    title = 'N061011 versus N61311',
    pCutoff = 10e-16,
    FCcutoff = 1.5,
    transcriptPointSize = 2.0,
    transcriptLabSize = 3.0,
    shape = c(1, 4, 23, 25),
    colAlpha = 1)

```

## Adjust cut-off lines and add extra threshold lines

The lines that are drawn to indicate cut-off points are also modifiable. The
parameter 'cutoffLineType' accepts the following values: "blank", "solid",
"dashed", "dotted", "dotdash", "longdash", and "twodash". The colour and
thickness of these can also be modified with 'cutoffLineCol' and
'cutoffLineWidth'. To disable the lines, set either cutoffLineType="blank"
or cutoffLineWidth=0.

Extra lines can also be added via 'hline' and 'vline' to display other cut-offs.

To make these more visible, we will also remove the default gridlines.

```{r ex5, fig.height = 8.5, fig.width = 7, fig.cap = "Adjust cut-off lines and add extra threshold lines."}

  EnhancedVolcano(res2,
    lab = rownames(res2),
    x = 'log2FoldChange',
    y = 'pvalue',
    xlim = c(-6, 6),
    title = 'N061011 versus N61311',
    pCutoff = 10e-12,
    FCcutoff = 1.5,
    transcriptPointSize = 1.5,
    transcriptLabSize = 3.0,
    colAlpha = 1,
    cutoffLineType = 'blank',
    cutoffLineCol = 'black',
    cutoffLineWidth = 0.8,
    hline = c(10e-12, 10e-36, 10e-60, 10e-84),
    hlineCol = c('grey0', 'grey25','grey50','grey75'),
    hlineType = 'longdash',
    hlineWidth = 0.8,
    gridlines.major = FALSE,
    gridlines.minor = FALSE)

```

## Adjust legend position, size, and text

The position of the legend can also be changed to "left" or "right"
(and stacked vertically), or 'top' or "bottom" (stacked horizontally). The
legend text, label size, and icon size can also be modified.

```{r ex6, fig.height = 8.5, fig.width = 12, fig.cap = "Adjust legend position, size, and text."}

  EnhancedVolcano(res2,
    lab = rownames(res2),
    x = 'log2FoldChange',
    y = 'pvalue',
    xlim = c(-6, 6),
    pCutoff = 10e-12,
    FCcutoff = 1.5,
    cutoffLineType = 'twodash',
    cutoffLineWidth = 0.8,
    transcriptPointSize = 3.0,
    transcriptLabSize = 4.0,
    colAlpha = 1,
    legend=c('NS','Log (base 2) fold-change','P value',
      'P value & Log (base 2) fold-change'),
    legendPosition = 'right',
    legendLabSize = 16,
    legendIconSize = 5.0)

```

Note: to make the legend completely invisible, specify:

```{r eval=FALSE}

legendVisible = FALSE

```

## Plot adjusted p-values

Volcano plots do not have to be produced with nominal (unadjusted P values),
even if this is the common practice. Simply provide a column name relating
to adjusted P values and you can also generate a volcano with these. In this
case, the cutoff for the P value then relates to the adjusted P value. Here,
we also modify the axis titles by supplying an expression via the bquote function.

```{r ex7, fig.height = 8.5, fig.width = 7, fig.cap = "Plot adjusted p-values."}

  EnhancedVolcano(res2,
    lab = rownames(res2),
    x = 'log2FoldChange',
    y = 'padj',
    xlim=c(-6,6),
    xlab = bquote(~Log[2]~ 'fold change'),
    ylab = bquote(~-Log[10]~adjusted~italic(P)),
    pCutoff = 0.0001,
    FCcutoff = 1.0,
    transcriptLabSize = 4.0,
    colAlpha = 1,
    legend=c('NS','Log2 FC','Adjusted p-value',
      'Adjusted p-value & Log2 FC'),
    legendPosition = 'bottom',
    legendLabSize = 10,
    legendIconSize = 3.0)

```

## Fit more labels by adding connectors

In order to maximise free space in the plot window, one can fit more
transcript labels by adding connectors from labels to points, where
appropriate. The width and colour of these connectors can also be
modified with 'widthConnectors' and 'colConnectors', respectively.
Further configuration is achievable via 'typeConnectors' ("open",
"closed"), 'endsConnectors' ("last", "first", "both"), and
lengthConnectors (default = unit(0.01, 'npc')).

The result may not always be desirable as it can make the plot
look overcrowded.

```{r ex8, fig.height = 8.5, fig.width = 11, fig.cap = "Fit more labels by adding connectors."}

  EnhancedVolcano(res2,
    lab = rownames(res2),
    x = 'log2FoldChange',
    y = 'pvalue',
    xlim = c(-6,6),
    xlab = bquote(~Log[2]~ 'fold change'),
    pCutoff = 10e-14,
    FCcutoff = 2.0,
    transcriptPointSize = 3.0,
    transcriptLabSize = 4.0,
    colAlpha = 1,
    legend=c('NS','Log (base 2) fold-change','P value',
      'P value & Log (base 2) fold-change'),
    legendPosition = 'right',
    legendLabSize = 12,
    legendIconSize = 4.0,
    drawConnectors = TRUE,
    widthConnectors = 0.2,
    colConnectors = 'grey30')

```

## Only label key transcripts

In many situations, people may only wish to label their key transcripts
/ transcripts of interest. One can therefore supply a vector of these
transcripts via the 'selectLab' parameter, the contents of which have
to also be present in the vector passed to 'lab'. In addition, only
those transcripts that pass both the cutoff for log2FC and P value will
be labelled.

```{r ex9, fig.height = 8.5, fig.width = 11, fig.cap = "Only label key transcripts."}

  EnhancedVolcano(res2,
    lab = rownames(res2),
    x = 'log2FoldChange',
    y = 'pvalue',
    selectLab = c('ENSG00000106565','ENSG00000187758'),
    xlim = c(-6,7),
    xlab = bquote(~Log[2]~ 'fold change'),
    pCutoff = 10e-14,
    FCcutoff = 2.0,
    transcriptPointSize = 3.0,
    transcriptLabSize = 5.0,
    shape = c(4, 35, 17, 18),
    colAlpha = 1,
    legend=c('NS','Log (base 2) fold-change','P value',
      'P value & Log (base 2) fold-change'),
    legendPosition = 'right',
    legendLabSize = 14,
    legendIconSize = 5.0)

```


## Draw labels in boxes

To improve label clarity, we can draw simple boxes around the plots
labels. This works much better when drawConnectors is also TRUE.

```{r ex10, fig.height = 8.5, fig.width = 12, fig.cap = "Draw labels in boxes."}

  EnhancedVolcano(res2,
    lab = rownames(res2),
    x = 'log2FoldChange',
    y = 'pvalue',
    selectLab = c('ENSG00000106565','ENSG00000187758',
      'ENSG00000230795', 'ENSG00000164530',
      'ENSG00000143153'),
    xlim = c(-5.5,8),
    xlab = bquote(~Log[2]~ 'fold change'),
    pCutoff = 10e-14,
    FCcutoff = 2.0,
    transcriptPointSize = 3.0,
    transcriptLabSize = 5.0,
    transcriptLabCol = 'black',
    transcriptLabFace = 'bold',
    boxedlabels = TRUE,
    colAlpha = 4/5,
    legend=c('NS','Log (base 2) fold-change','P value',
      'P value & Log (base 2) fold-change'),
    legendPosition = 'right',
    legendLabSize = 14,
    legendIconSize = 4.0,
    drawConnectors = TRUE,
    widthConnectors = 1.0,
    colConnectors = 'black')

```


## Over-ride colouring scheme with custom key-value pairs

In certain situations, one may wish to over-ride the default colour scheme
with their own colour-scheme, such as colouring transcripts by pathway,
cell-type or group. This can be achieved by supplying a named vector as
'colCustom'.

In this example, we just wish to colour all transcripts with log2FC > 2.5 as
'high' and those with log2FC < -2.5 as 'low'.

```{r eval = TRUE}

  # create custom key-value pairs for 'high', 'low', 'mid' expression by fold-change
    # set the base colour as 'black'
    keyvals <- rep('black', nrow(res2))

    # set the base name/label as 'Mid'
    names(keyvals) <- rep('Mid', nrow(res2))

    # modify keyvals for transcripts with fold change > 2.5
    keyvals[which(res2$log2FoldChange > 2.5)] <- 'gold'
    names(keyvals)[which(res2$log2FoldChange > 2.5)] <- 'high'

    # modify keyvals for transcripts with fold change < -2.5
    keyvals[which(res2$log2FoldChange < -2.5)] <- 'royalblue'
    names(keyvals)[which(res2$log2FoldChange < -2.5)] <- 'low'

    unique(names(keyvals))
    unique(keyvals)
    keyvals[1:20]

```

```{r ex11, fig.height = 9.5, fig.width = 15, fig.cap = "Over-ride colouring scheme with custom key-value pairs."}

  p1 <- EnhancedVolcano(res2,
    lab = rownames(res2),
    x = 'log2FoldChange',
    y = 'pvalue',
    selectLab = rownames(res2)[which(names(keyvals) %in% c('high', 'low'))],
    xlim = c(-6.5,6.5),
    xlab = bquote(~Log[2]~ 'fold change'),
    title = 'Custom colour over-ride',
    pCutoff = 10e-14,
    FCcutoff = 1.0,
    transcriptPointSize = 3.5,
    transcriptLabSize = 4.5,
    shape = c(6, 4, 2, 11),
    colCustom = keyvals,
    colAlpha = 1,
    legendPosition = 'top',
    legendLabSize = 15,
    legendIconSize = 5.0,
    drawConnectors = TRUE,
    widthConnectors = 0.5,
    colConnectors = 'grey50',
    gridlines.major = TRUE,
    gridlines.minor = FALSE,
    border = 'partial',
    borderWidth = 1.5,
    borderColour = 'black')

  p2 <- EnhancedVolcano(res2,
    lab = rownames(res2),
    x = 'log2FoldChange',
    y = 'pvalue',
    selectLab = rownames(res2)[which(names(keyvals) %in% c('high', 'low'))],
    xlim = c(-6.5,6.5),
    xlab = bquote(~Log[2]~ 'fold change'),
    title = 'No custom colour over-ride',
    pCutoff = 10e-14,
    FCcutoff = 1.0,
    transcriptPointSize = 3.5,
    transcriptLabSize = 4.5,
    colCustom = NULL,
    colAlpha = 1,
    legendPosition = 'top',
    legendLabSize = 15,
    legendIconSize = 5.0,
    drawConnectors = FALSE,
    widthConnectors = 0.5,
    colConnectors = 'grey50',
    gridlines.major = TRUE,
    gridlines.minor = FALSE,
    border = 'full',
    borderWidth = 1.0,
    borderColour = 'black')

  library(gridExtra)
  library(grid)
  grid.arrange(p1, p2,
    ncol=2,
    top = textGrob('EnhancedVolcano',
      just = c('center'),
      gp = gpar(fontsize = 32)))
  grid.rect(gp=gpar(fill=NA))

```


## Over-ride colour and/or shape scheme with custom key-value pairs

In this example, we first over-ride the existing shape scheme and then
both the colour and shape scheme at the same time.


```{r eval = TRUE}

  # define different cell-types that will be shaded
  celltype1 <- c('ENSG00000106565', 'ENSG00000002933',
    'ENSG00000165246', 'ENSG00000224114')
  celltype2 <- c('ENSG00000230795', 'ENSG00000164530',
    'ENSG00000143153', 'ENSG00000169851',
    'ENSG00000231924', 'ENSG00000145681')

  # create custom key-value pairs for different cell-types
    # set the base shape as '3'
    keyvals.shape <- rep(3, nrow(res2))

    # set the base name/label as 'PBC'
    names(keyvals.shape) <- rep('PBC', nrow(res2))

    # modify the keyvals for cell-type 1
    keyvals.shape[which(rownames(res2) %in% celltype1)] <- 17
    names(keyvals.shape)[which(rownames(res2) %in% celltype1)] <- 'Cell-type 1'

    # modify the keyvals for cell-type 2
    keyvals.shape[which(rownames(res2) %in% celltype2)] <- 64
    names(keyvals.shape)[which(rownames(res2) %in% celltype2)] <- 'Cell-type 2'

    unique(names(keyvals.shape))
    unique(keyvals.shape)
    keyvals.shape[1:20]

```

```{r ex12, fig.height = 9, fig.width = 17, fig.cap = "Over-ride colour and/or shape scheme with custom key-value pairs."}

  p1 <- EnhancedVolcano(res2,
    lab = rownames(res2),
    x = 'log2FoldChange',
    y = 'pvalue',
    selectLab = rownames(res2)[which(names(keyvals) %in% c('high', 'low'))],
    xlim = c(-6.5,6.5),
    xlab = bquote(~Log[2]~ 'fold change'),
    title = 'Custom shape over-ride',
    pCutoff = 10e-14,
    FCcutoff = 1.0,
    transcriptPointSize = 3.5,
    transcriptLabSize = 4.5,
    shapeCustom = keyvals.shape,
    colCustom = NULL,
    colAlpha = 1,
    legendLabSize = 15,
    legendPosition = 'left',
    legendIconSize = 5.0,
    drawConnectors = TRUE,
    widthConnectors = 0.5,
    colConnectors = 'grey50',
    gridlines.major = TRUE,
    gridlines.minor = FALSE,
    border = 'partial',
    borderWidth = 1.5,
    borderColour = 'black')

  # create custom key-value pairs for 'high', 'low', 'mid' expression by fold-change
    # set the base colour as 'black'
    keyvals.colour <- rep('black', nrow(res2))

    # set the base name/label as 'Mid'
    names(keyvals.colour) <- rep('Mid', nrow(res2))

    # modify keyvals for transcripts with fold change > 2.5
    keyvals.colour[which(res2$log2FoldChange > 2.5)] <- 'gold'
    names(keyvals.colour)[which(res2$log2FoldChange > 2.5)] <- 'high'

    # modify keyvals for transcripts with fold change < -2.5
    keyvals.colour[which(res2$log2FoldChange < -2.5)] <- 'royalblue'
    names(keyvals.colour)[which(res2$log2FoldChange < -2.5)] <- 'low'

    unique(names(keyvals.colour))
    unique(keyvals.colour)

  p2 <- EnhancedVolcano(res2,
    lab = rownames(res2),
    x = 'log2FoldChange',
    y = 'pvalue',
    selectLab = rownames(res2)[which(names(keyvals) %in% c('High', 'Low'))],
    xlim = c(-6.5,6.5),
    xlab = bquote(~Log[2]~ 'fold change'),
    title = 'Custom shape & colour over-ride',
    pCutoff = 10e-14,
    FCcutoff = 1.0,
    transcriptPointSize = 5.5,
    transcriptLabSize = 0.0,
    shapeCustom = keyvals.shape,
    colCustom = keyvals.colour,
    colAlpha = 1,
    legendPosition = 'top',
    legendLabSize = 15,
    legendIconSize = 5.0,
    drawConnectors = TRUE,
    widthConnectors = 0.5,
    colConnectors = 'grey50',
    gridlines.major = TRUE,
    gridlines.minor = FALSE,
    border = 'full',
    borderWidth = 1.0,
    borderColour = 'black')

  library(gridExtra)
  library(grid)
  grid.arrange(p1, p2,
    ncol=2,
    top = textGrob('EnhancedVolcano',
      just = c('center'),
      gp = gpar(fontsize = 32)))
  grid.rect(gp=gpar(fill=NA))

```

## Shade certain transcripts

In this example we add an extra level of highlighting key transcripts by
shading.

This feature works best for shading just 1 or 2 key transcripts. It is
expected that the user can use the 'shapeCustom' parameter for more in depth
identification of different types of transcripts.

```{r eval = TRUE}

  # define different cell-types that will be shaded
  celltype1 <- c('ENSG00000106565', 'ENSG00000002933')
  celltype2 <- c('ENSG00000230795', 'ENSG00000164530')

```

```{r ex13, fig.height = 9, fig.width = 16, fig.cap = "Shade certain transcripts."}

  p1 <- EnhancedVolcano(res2,
    lab = rownames(res2),
    x = 'log2FoldChange',
    y = 'pvalue',
    selectLab = celltype1,
    xlim = c(-6.5,6.5),
    xlab = bquote(~Log[2]~ 'fold change'),
    title = 'Shading cell-type 1',
    pCutoff = 10e-14,
    FCcutoff = 1.0,
    transcriptPointSize = 8.0,
    transcriptLabSize = 5.0,
    transcriptLabCol = 'purple',
    transcriptLabFace = 'bold',
    boxedlabels = TRUE,
    shape = 42,
    colCustom = keyvals,
    colAlpha = 1,
    legendPosition = 'top',
    legendLabSize = 15,
    legendIconSize = 5.0,
    shade = celltype1,
    shadeLabel = 'Cell-type I',
    shadeAlpha = 1/2,
    shadeFill = 'purple',
    shadeSize = 1,
    shadeBins = 5,
    drawConnectors = TRUE,
    widthConnectors = 1.0,
    colConnectors = 'grey30',
    gridlines.major = TRUE,
    gridlines.minor = FALSE,
    border = 'partial',
    borderWidth = 1.5,
    borderColour = 'black')

  p2 <- EnhancedVolcano(res2,
    lab = rownames(res2),
    x = 'log2FoldChange',
    y = 'pvalue',
    selectLab = celltype2,
    xlim = c(-6.5,6.5),
    xlab = bquote(~Log[2]~ 'fold change'),
    title = 'Shading cell-type 2',
    pCutoff = 10e-14,
    FCcutoff = 1.0,
    transcriptLabSize = 5.0,
    transcriptLabCol = 'forestgreen',
    transcriptLabFace = 'bold',
    shapeCustom = keyvals.shape,
    colCustom = keyvals.colour,
    colAlpha = 1,
    legendPosition = 'top',
    transcriptPointSize = 4.0,
    legendLabSize = 15,
    legendIconSize = 5.0,
    shade = celltype2,
    shadeLabel = 'Cell-type II',
    shadeAlpha = 1/2,
    shadeFill = 'forestgreen',
    shadeSize = 1,
    shadeBins = 5,
    drawConnectors = TRUE,
    widthConnectors = 1.0,
    colConnectors = 'grey30',
    gridlines.major = TRUE,
    gridlines.minor = FALSE,
    border = 'full',
    borderWidth = 1.0,
    borderColour = 'black')

  library(gridExtra)
  library(grid)
  grid.arrange(p1, p2,
    ncol=2,
    top = textGrob('EnhancedVolcano',
      just = c('center'),
      gp = gpar(fontsize = 32)))
  grid.rect(gp=gpar(fill=NA))

```

# Acknowledgments

The development of *EnhancedVolcano* has benefited from contributions
and suggestions from:

Sharmila Rana

[Myles Lewis](https://www.qmul.ac.uk/whri/people/academic-staff/items/lewismyles.html)

Luke Dow - Assistant Professor at Weill Cornell Medicine

Tokhir Dadaev - Institute of Cancer Research

Alina Frolova

Venu Thatikonda (Deutsches Krebsforschungszentrum (DKFZ) / German Cancer Research Center)

David Wheeler - Montana State University

David Kulp

# Session info

```{r}

sessionInfo()

```

# References

